---
title: "PPHA 34600 - Problem Set 2 Solutions"
author: "Jocelyn Sánchez, Daniela Avayú, Luis E. Santos Burgoa"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    fig_width: 7
    fig_height: 4.5
header-includes:
  - \usepackage{amsmath}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

data <- read.csv("data/ps2_data_spring26.csv")
names(data) <- c(
  "owner_older40",
  "owner_male",
  "num_livestock",
  "years_edu",
  "num_employees",
  "grid_subsidy_amount",
  "grid_connection",
  "firm_revenue"
)
```

# Question 1:

PEDAP are interested in answering the following question: What is the average effect of grid electricity connection on firm annual revenues (measured in USD)? To make sure everybody is on the same page, explain to them what the ideal experiment would be for answering this question. Next, use math (the potential outcomes framework) and words to explain what you would estimate and how you would do so. Make sure to be clear about the unit of analysis (i.e., what is “i” here?). Word limit: 250 words

The ideal experiment is a randomized controlled trial (RCT). We would randomly assign grid electricity connections to firms, with some receiving connections (treatment group) and others remaining unconnected (control group). Randomization ensures that any differences between groups are causal, not due to pre-existing differences. We would then measure annual firm revenue for both groups and compare the difference.

# Computing ATE

```{r}
ate_simple <- mean(data$firm_revenue[data$grid_connection == 1]) - 
              mean(data$firm_revenue[data$grid_connection == 0])
ate_simple
```

# Question 2:

PEDAP has some data available, and they believe they can compare firms that are connected to the grid, to those who aren’t. They make a dataset available to you, in theform of ps3_data_Spring26.csv. Explain what you would recover if you simply compared connected to non-connected firms.

Produce a balance table which displays the differences on observable characteristics between firms that are connected to the grid versus those that are not connected. Interpret this table. Does this table give you more or less faith in the comparison PEDAP would like to make? Word limit: 250 words.

```{r}
# Balance Table

characteristics <- c("owner_older40", "owner_male", "num_livestock", "years_edu", "num_employees", "grid_subsidy_amount")

balance_table <- data.frame()

for (var in characteristics) {
  # Means
  mean_connected <- mean(data[[var]][data$grid_connection == 1], na.rm = TRUE)
  mean_unconnected <- mean(data[[var]][data$grid_connection == 0], na.rm = TRUE)
  difference <- mean_connected - mean_unconnected
  
  # T-test
  t_test <- t.test(data[[var]][data$grid_connection == 1], 
                   data[[var]][data$grid_connection == 0])
  p_value <- t_test$p.value
  
  # Significance stars
  sig <- ifelse(p_value < 0.01, "***", 
                ifelse(p_value < 0.05, "**", 
                       ifelse(p_value < 0.10, "*", "")))
  
  balance_table <- rbind(balance_table, data.frame(
    Variable = var,
    Connected = round(mean_connected, 4),
    Not_Connected = round(mean_unconnected, 4),
    Difference = round(difference, 4),
    P_value = round(p_value, 4),
    Significant = sig
  ))
}

```

balance_table

The balance table shows that firms with grid connections differ systematically from those without, indicating that treatment is not randomly assigned. Connected firms are much more likely to have older owners, male owners, and more educated owners, and they also tend to receive higher subsidies and have slightly more employees; all of these differences are statistically significant, except for livestock ownership, which appears balanced across groups. This pattern suggests clear selection bias, meaning that firms receiving grid connections were already more advantaged along several dimensions before treatment. As a result, the simple average treatment effect estimate of 12,992 is likely biased upward and does not reflect a purely causal impact of grid access on revenue. To obtain a more credible estimate, it is necessary to control for these pre-existing differences using a regression framework that adjusts for observable characteristics.

# Question 3:

(Optional, ungraded question) PEDAP acknowledges that a simple comparison may not work, but they still want an estimate of the impacts of being connected to the grid. PEDAP’s CEO recently took a class on impact evaluation and learned about selection-on-observables designs. They have asked you to apply this methodology. Use a regression-based approach to estimate the effect of the grid-connection on firm revenue. Interpret your results. What are the strengths and weaknesses of this approach? How do your results differ from what you find if you instead use the naive estimator? Word limit: 250 words.

```{r}
# Regression-based Selection-on-Observables Estimate
# Estimate treatment effect controlling for observable characteristics
model_controls <- lm(
  firm_revenue ~ grid_connection +
    owner_older40 +
    owner_male +
    num_livestock +
    years_edu +
    num_employees +
    grid_subsidy_amount,
  data = data
)

# Regression
summary(model_controls)

# Extract treatment effect estimate
ate_adjusted <- coef(model_controls)["grid_connection"]
ate_adjusted

# Difference between naive and adjusted estimates
difference <- ate_simple - ate_adjusted
difference

# Comparison table
comparison <- data.frame(
  Estimator = c("Naive Difference in Means", "Regression Adjusted"),
  Estimate = c(ate_simple, ate_adjusted)
)

print(comparison)
```


# Question 4:

Use an exact matching approach to estimate the effect of a grid connection on firm revenue. What variables did you decide to include in the matching procedure (remember you don’t have to use all the variables at your disposal)? Interpret your results. What are the strengths and weaknesses of this approach? How do your results differ from what you find if you use the naive estimator? Did you run into the Curse of Dimensionality with this analysis? If yes, describe how it affected your approach. If not, describe how the Curse could have generated problems in this setting. Word limit: 250 words.

```{r}
library(MatchIt)
library(dplyr)

# Exact matching: covariates chosen for matching wil be education, gender and firm size

m.out <- matchit(
  grid_connection ~ years_edu + owner_male + num_employees,
  data = data,
  method = "exact"
)

# Extract matched dataset
matched_data <- match.data(m.out)

# Weighted Matched ATE
ate_weighted <- matched_data %>%
  group_by(subclass) %>%
  summarise(
    diff = mean(firm_revenue[grid_connection == 1]) -
           mean(firm_revenue[grid_connection == 0]),
    n = n()
  ) %>%
  summarise(weighted_ate = weighted.mean(diff, n)) %>%
  pull(weighted_ate)

# Comparison
results <- data.frame(
  Method = c("Naive", "Weighted Matching"),
  Estimate = c(ate_simple, ate_weighted)
)

print(results)

```
We implemented an exact matching approach by matching firms with and without grid connection on a selected set of pre-treatment covariates (owner education, owner gender, firm size) as matching variables, since these are strongly related to both the likelihood of receiving a grid connection and firm revenue, while keeping the dimensionality manageable. 

The main strength of exact matching is that it creates highly comparable treatment and control groups, improving internal validity by eliminating bias from the selected covariates. However, its key weakness is data loss: many observations cannot be matched exactly when multiple covariates are used, reducing sample size and potentially affecting external validity. It also does not adjust for unobserved confounders.

Regarding the Curse of Dimensionality,if more covariates had been included, finding exact matches would become increasingly difficult, potentially leaving very few or no matched pairs.

The resulting estimated effect of grid connection is positive but smaller than the naive estimator of 12,992, suggesting that part of the raw difference was driven by observable differences between firms rather than the treatment itself.

# Question 5:

PEDAP has decided they are not interested in using a selection-on-observables approach after your discussions with them. Instead, they mention that they have also heard about a strategy called instrumental variables (IV) and are considering using land gradient as an instrument. They would like you to discuss the plausibility of this instrument and outline its limitations. They would also like you to propose an alternative instrumental variable that could be used to evaluate the effect of grid electricity connection on firm revenues. Why would your proposed instrument be a good one? Do you have any concerns about your ability to estimate the treatment effect using your instrument? If so, explain why; if not, explain why not. Word limit: 250 words.

Using land gradient as an instrument for grid connection is plausible in this context because it is likely correlated with the cost of extending electricity infrastructure: steeper terrain makes grid expansion more difficult and expensive, reducing the probability that a firm is connected. This supports the relevance condition. However, its main limitation concerns the exclusion restriction: land gradient may directly affect firm revenue through channels unrelated to electricity access, such as transportation costs, agricultural productivity, or market access. If this is the case, the instrument would be invalid because it would not affect revenue only through grid connection.

An alternative instrumental variable could be the eligibility for a government program that reduces or fully waives grid connection fees, where access to the subsidy is determined through a lottery-based selection within eligible areas. This instrument would be appropriate because it is strongly related to the likelihood of receiving a grid connection, as lowering the cost of connection directly increases the probability that firms adopt electricity access. At the same time, if assignment to the subsidy is random, it can plausibly satisfy the exclusion restriction, meaning it affects firm revenues only through its impact on grid connection rather than through other channels.

However, there are some concerns with this approach. First, the exclusion restriction may still be threatened if the subsidy program affects firms through additional mechanisms, such as complementary investments, increased local government attention, or other development programs bundled with electrification efforts. Second, even if assignment is randomized, spillovers may occur if nearby firms also benefit from improved infrastructure or increased economic activity, which could bias estimates. 
